\bf Estimating the number of clusters using cross-validation

### 2018-03-09 cvclust.tex
more progress

In unsupervised learning problems like factor analysis and clustering, the
features of the observations are not naturally partitioned into ``predictors''
and ``responses'', so we cannot directly apply the cross-validation procedure
described above.  For factor analysis, there are at least two versions of
cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
in each fold we leave out a subset of the elements of the data matrix.  Wold's
procedure works well empirically, but does not have any theoretical support,
and it requires a factor analysis procedure that can handle missing data.
\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
each fold designates a subset of the data matrix columns to be response and a
subset of the rows to be test data.  This generalized a procedure due to
\citet{gabriel2002biblot}, who proposed holding out a single column and a
single row at each fold. Gabriel's cross validation is analogous to
leave-one-out cross-validation, while Owen and Perry's generalization handles
more general leave-outs.

@@

Our version of Gabriel cross validation for clustering works by performing a
sequence of ``folds'' over the data. We use these folds to estimate a version
of prediction error (cross-validation error) for each possible value of $k$;
we then choose the value $\hat k$ with the smallest cross-validation error.

@@

We have stated and prooved Proposition~\ref{prop:twoclust} for data in $d = 2$
dimensions, but there is nothing essential about the particulars of the
dimension; we generalize the result to arbitrary dimension in
Appendex~\ref{app:twoclust-anyd}.

@@

Our correlation correction is not backed by a rigorous theoretical
justification. Further, since our initial estimate $\hat k_0$ tends to
over-estimate the true number of clusters, our covariance estimate $\hat
\Sigma$ will often be biased. Fortunately, we do not need to estimate $\Sigma$
with high accuracy. For our correction to be useful, we only need $\hat
\Sigma^{-1/2}$ to have lower within-cluster correlation than the raw data. In
our simulations and in our empirical validation study in
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation} we
demonstrate that despite its theoretical shortcoming, our ad-hoc correlation
adjustment often performs well in practice.

@@

In all simulation settings, we randomly generate cluster centers by drawing
from a multivariate normal distribution with covariance matrix $\tau I$,
conditional on the cluster centers being well-separated (if the distance
between any two cluster centers is less than $1$, then we re-draw a new set of
cluster centers). We choose $\tau$ to make the probability the cluster
centers being well-separated on the first draw to be equal to approximately
50\%. Many of our simulation settings are chosen to mimic the settings used by
\cite{tibshirani2001estimating}.

@@

The first data set is congressional voting data consisting of voting records
of the second session of the $98$th United States Congress,
\citep{schlimmer1987concept}.  This data set includes votes for legislators on
the $P = 16$ key votes.  For each vote, each legislator either votes
positively (``yea'') or negatively (``nay''). We removed legislators with missing
votes.  This results in $N = 232$ remaining records, with $124$ democrat and
$108$ republican. There are $k = 2$ clusters of legislators, corresponding to
political party.

@@

In this paper, we proposed a new approach to estimate the number of clusters
to be used in $k$-means clustering. The intuition behind our proposed method
is to transform the unsupervised learning problem into a supervised learning
problem via a form of Gabriel cross validation.  We proved that our method is
self-consistent, and we analyzed its behavior in some special cases of
Gaussian mixture models.  Using simulations and real data examples, we
demonstrated that our method has good performance, competitive with existing
approaches. The simulations and empirical benchmarks demonstrate the
advantages of our method. In the yeast cell cycle application, our method
was able to identify meaningful gene groups with a small number of clusters.

>>>>

In unsupervised learning problems like factor analysis and clustering, the
features of the observations are not naturally partitioned into ``predictors''
and ``responses'', so we cannot directly apply the cross-validation procedure
described above.  For factor analysis, there are at least two versions of
cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
in each fold we leave out a subset of the elements of the data matrix.  Wold's
procedure works well empirically, but does not have any theoretical support,
and it requires a factor analysis procedure that can handle missing data.
\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
each fold designates a subset of the data matrix columns to be response and a
subset of the rows to be test data.  This generalized a procedure due to
\citet{gabriel2002biblot}, who proposed holding out a single column and a
single row at each fold. Gabriel's cross-validation is analogous to
leave-one-out cross-validation, while Owen and Perry's generalization handles
more general leave-outs.

@@

Our version of Gabriel cross-validation for clustering works by performing a
sequence of ``folds'' over the data. We use these folds to estimate a version
of prediction error (cross-validation error) for each possible value of $k$;
we then choose the value $\hat k$ with the smallest cross-validation error.

@@

We have stated and proved Proposition~\ref{prop:twoclust} for data in $d = 2$
dimensions, but there is nothing essential about the particulars of the
dimension; we generalize the result to arbitrary dimension in
Appendex~\ref{app:twoclust-anyd}.

@@

Our correlation correction is not backed by a rigorous theoretical
justification. Further, since our initial estimate $\hat k_0$ tends to
over-estimate the true number of clusters in the presence of strong
correlation, our covariance estimate $\hat \Sigma$ will often be biased.
Fortunately, we do not need to estimate $\Sigma$ with high accuracy. For our
correction to be useful, we only need $\tilde{\dataX}$ to have lower
within-cluster correlation than the raw data~$\dataX$. In our simulations and
in our empirical validation study in Sections~\ref{sec:simulations}
and~\ref{sec:empirical-validation} we demonstrate that despite its theoretical
shortcoming, our ad-hoc correlation adjustment often performs well in
practice.

@@

In all simulation settings, we randomly generate cluster centers by drawing
from a multivariate normal distribution with covariance matrix $\tau I$,
conditional on the cluster points being well-separated. By ``well-separated''
we mean that the distance from any point to its own cluster center is less
than the distance of that point to any other cluster center by a margin of
at least $1.0$ units. In any replicate where the clusters are not
well-separated, we discard the replicate and generate another set of points.
We choose $\tau$ to make the probability the cluster
centers being well-separated on the first draw to be equal to approximately
50\%.  The ``well-separated'' condition and some of our simulation settings
are chosen to mimic those used by~\cite{tibshirani2001estimating} in
settings (c) and (d) of their Section~6.

@@

The first data set is congressional voting data consisting of voting records
of the second session of the $98$th United States Congress
\citep{schlimmer1987concept}.  This data set includes votes for legislators on
the $P = 16$ key votes.  For each vote, each legislator either votes
positively (``yea'') or negatively (``nay''). We removed legislators with missing
votes.  This results in $N = 232$ remaining records, with $124$ democrat and
$108$ republican. There are $k = 2$ clusters of legislators, corresponding to
political party.

@@

In this paper, we proposed a new approach to estimate the number of clusters
to be used in $k$-means clustering. The intuition behind our proposed method
is to transform the unsupervised learning problem into a supervised learning
problem via a form of Gabriel cross-validation.  We proved that our method is
self-consistent, and we analyzed its behavior in some special cases of
Gaussian mixture models.  Using simulations and real data examples, we
demonstrated that our method has good performance, competitive with existing
approaches. The simulations and empirical benchmarks demonstrate the
advantages of our method. In the yeast cell cycle application, our method
was able to identify meaningful gene groups with a small number of clusters.
